The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-06 thesis/extra_density.tex
thesis; julio oct 4 comments

In this section, we describe the details of J+24's likelihoods. We then
test the reliability of the derived density profiles for Scl and UMi,
comparing methodologies, samples, and to literature. Finally, we present
a comparison of J+24's samples to a non-parametric Bayesian-derived
density profile, finding similar results except in the very outer
regions. Past a ``limiting radius'', the density profiles become
susceptible to the likelihood specification and may select mostly
background stars. In conclusion, the density profiles from J+24 converge
among methods out to our limiting radii derived here.

@@

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
  quality astrometry. \texttt{ruwe} is a measure of the excess
  astrometric noise on fitting a consistent parallax-proper motion
  solution \citep[see][]{lindegren+2021}.\\
\item
  3\(\sigma\) consistency of measured parallax with dwarf's distance,
  with \citet{lindegren+2021}'s zero-point correction. Note that the
  typical satellite parallax is very small.
\item
  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
  10\(\,{\rm mas\ yr^{-1}}\). This corresponds to tangental velocities
  of \(\gtrsim 500\) km/s at distances larger than 10 kpc.
\item
  Corrected colour excess is within 3\(\sigma\) of the expected
  distribution from \citet{riello+2021}. Removes stars with unreliable
  photometry.
\item
  De-reddened \(G\) magnitude is between
  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
  and stars significantly brighter than the tip of the red giant branch
  (TRGB) magnitude plus the distance modulus uncertainty
  \(\sigma_{\rm DM}\).
\item
  De-reddened colour is between \(-0.5 < {\rm BP - RP} <  2.5\). Removes
  stars substantially outside the expected CMD.
\end{itemize}

@@

The background likelihoods are instead empirically constructed. Stars
stars outside of 5\(R_h\) passing the quality cuts estimate the
background density in PM and CMD space. The density is a sum of
bivariate gaussians with variances based on Gaia uncertainties (and
covariance for proper motions). The spatial background likelihood is
assumed to be constant.

@@

\section{Alternative methodologies}\label{alternative-methodologies}

\subsection{Simplistic cuts}\label{simplistic-cuts}

Besides a global background subtraction (as used in
Section~\ref{sec:observations}), absolute cuts in CMD and PM represent
perhaps the next simplest method to construct density profiles. We
create cuts using only the following

\begin{itemize}
\tightlist
\item
  (BP-RP, G) within the vertices defined in Table~\ref{tbl:colour_cuts}
\item
  PM within a \(1\masyr\) radius of the galaxies measured value.
\item
  Parallax within \(3\sigma\) of the dwarf distance
\item
  \texttt{ruwe} \textless{} 1.3
\end{itemize}

\begin{table*}[t]
\centering
\caption[Colour cuts for density profiles.]{Colour cuts used in samples in this section. Each pair of columns contains the vertices of a polygon in the CMD within which we select stars to derive density profiles. }
\label{tbl:colour_cuts}
\begin{tabular}{llllllll}
\toprule
Scl &  &  &  & UMi &  &  & \\
\midrule
Gaia &  & DELVE &  & Gaia &  & Unions & \\
BP-RP & G & g-r & g & BP-RP & G & r-i & r\\
-0.21 & 20.58 & 0.32 & 22.58 & 0.49 & 20.83 & -0.485 & 24.13\\
-0.1 & 20.1 & 0.51 & 23.07 & 1.24 & 20.9 & 0.145 & 23.11\\
0.28 & 19.77 & 0.56 & 21.49 & 1.18 & 19.65 & 0.201 & 21.41\\
0.7 & 19.64 & 0.71 & 19.7 & 1.23 & 18.12 & 0.21 & 20.05\\
1.02 & 18.5 & 0.96 & 18.49 & 1.39 & 17.09 & 0.288 & 17.23\\
1.14 & 17.59 & 0.84 & 18.35 & 1.32 & 16.94 & 0.226 & 17.21\\
1.41 & 16.73 & 0.6 & 19.65 & 0.96 & 18.58 & 0.151 & 19.3\\
1.68 & 15.94 & 0.42 & 21.42 & 0.88 & 19.22 & 0.145 & 19.75\\
1.94 & 15.94 &  &  & 0.4 & 19.52 & 0.026 & 19.86\\
1.91 & 16.8 &  &  & -0.03 & 19.78 & 0.011 & 19.52\\
1.68 & 17.01 &  &  & 0.06 & 20.16 & -0.155 & 19.95\\
1.45 & 17.87 &  &  & 0.83 & 19.72 & -0.22 & 20.34\\
1.32 & 19.16 &  &  & 0.21 & 20.83 & -0.176 & 20.47\\
1.26 & 20.1 &  &  &  &  & -0.098 & 20.33\\
1.35 & 21 &  &  &  &  & 0.008 & 19.84\\
0.46 & 21 &  &  &  &  & 0.142 & 19.75\\
0.82 & 20.27 &  &  &  &  & 0.098 & 21.45\\
0.39 & 20.33 &  &  &  &  & -0.052 & 22.38\\
0.16 & 20.76 &  &  &  &  & -0.388 & 23.42\\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Alternative samples}\label{alternative-samples}

For the DELVE sample \citep{drlica-wagner+2022}, we select everything in
data release 2 between RA of 11 and 19, and declination of -37.7 and
-29.7.

We also consider samples from the Ultraviolet Near-Infrared Optical
Northern Survey \citep[UNIONS][]{gwyn+2025} for Ursa Minor. We select
stars within a 5 degree tangent square of UMi with no
\texttt{FLAGS\_CFIS} set and \texttt{s21} and \texttt{s31} both \(<3\)
(not extended sources).

\subsection{A Bayesian histogram
model}\label{a-bayesian-histogram-model}

To address the concerns discussed above, we consider here a
non-parametric model to fit the density in each bin. This model extends
the J+24 framework with two notable differences. (1) The systemic proper
motions are held fixed. This allows for a much more efficient
calculation of the likelihoods. (2) The spatial likelihood for the
satellite is a piecewise constant density function.While this model
theoretically has many more parameters to fit (one for each bin), each
parameter is independently estimated based on only the stars in that
specific bin.

@@

In detail, we parameterize \(f_{\rm sat}\) in terms of the log-relative
satellite density. The model construction is then \begin{equation}{
\theta_i \equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}})
}\end{equation}

\begin{equation}{
f_{\rm sat} = \frac{10^{\theta_i}}{1 + 10^{\theta_i}}
}\end{equation}

\begin{equation}{
\theta_i \sim {\rm Uniform}(-12, 6)
}\end{equation}

\begin{equation}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm CMD, PM} + (1-f_{\rm sat}){\cal L}_{\rm CMD, PM}
}\end{equation}

For each satellite, we hold the structural parameters fixed to those in
J+24, create bins which are the larger of the bin containing 20 stars or
0.05dex in \(\log R_{\rm ell}\). We then sample each chain 1000 steps
with 48 realizations using Turing.jl and the No-U Turns Sampler. The
chains appear to converge excellently.

@@

\section{Comparison of density profiles}\label{sec:density_extra}

In this section, we discuss additional tests and verification of the
derived density profiles. In particular, we check that methodology
(simpler cuts, circularized radii, algorithm version) do not
substantially affect the density profile. We also compile density
profiles presented in the literature as reference. In all cases, the
density profiles appear to have excellent convergence out to
\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
where the background dominates.

\textbf{Spatial likelihood.} J+24 method was designed in particular to
detect the presence of a density excess and find individual stars at
large radii to be followed up. We are more interested in accurately
quantifying the density profile and size of any perturbations. One
potential problem with using J+24's candidate members is that the
algorithm assumes the density is either described by a single or double
exponential. If this model does not accurately match the actual density
profile of the dwarf galaxy, we want to understand the impact of this
assumption.In particular, in Fig.~\ref{fig:scl_observed_profiles},
notice that the \(P_{\rm sat}\) selection method produces small
errorbars, even when the density is more than 1 dex below the local
background. These stars are likely selecting stars from the statistical
MW background consistent with UMi PM+CMD, recovering the assumed density
profile. As a result, the reliability of these density profiles below
the CMD+PM background may be questionable. A more robust analysis,
removing this particular density assumption, would be required to more
appropriately represent the knowledge of the density profile as the
background begins to dominate.

\textbf{\emph{Gaia} systematics}. While \emph{Gaia} has shown excellent
performance, some notable limitations may introduce problems in our
interpretation and reliability of density profiles. Gaia systematics in
proper motions and parallaxes are typically smaller than the values for
sources of magnitudes \(G\in[18,20]\). Since we use proper motions and
parallaxes as general consistency with the dwarf, and factor in
systematic uncertainties in each case, these effects should not be too
significant. However, the systematic proper motion uncertainties becomes
the dominant source of uncertainty in the derived systemic proper
motions of each galaxy (see
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).

\textbf{Completeness}. \emph{Gaia} shows high but imperfect
completeness, particularly showing limitations in crowded fields and for
faint sources (\(G\gtrapprox20\)). As discussed in
\citet{fabricius+2021}, for the high stellar densities in globular
clusters, the completeness relative to HST varies significantly with the
stellar density. However, the typical stellar densities of dwarf
galaxies are much lower, at about 20 stars/arcmin = 90,000 stars /
degree, lower than the lowest globular cluster densities and safely
below the crowding limit of 750,000 objects/degree for BP/RP photometry.
In \citet{fabricius+2021}, for the lowest density globular clusters, the
completeness down to \(G\approx 20\) is \(\sim 80\%\). Closely separated
stars pose problems for Gaia's on-board processing, as the pixel size is
59x177 mas on the sky. This results in a reduction of stars separated by
less than 1.5'' and especially for stars separated by less than 0.6
arcseconds. The astrometric parameters of closely separated stars
furthermore tends to be of lower quality \citep{fabricius+2021}.
However, even for the denser field of Fornax, only about 3\% of stars
have a neighbour within 2 arc seconds, so multiplicity should not affect
completeness too much (except for unresolved binaries). One potential
issue is that the previous analyses do not account for our cuts on
quality and number of astrometric parameters. These could worsen
completeness, particularly since the BP-RP spectra are more sensitive to
dense fields.

In Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras} we
show the results of limiting the density profile to the brightest half
of stars. We find no substantive differences. If \emph{Gaia}'s
incompleteness is inhomogeneous, then the variations in \emph{Gaia}'s
completeness do not strongly affect the density profiles.

\textbf{Structural uncertainties}. J+24 do not account for structural
uncertainties in dwarfs for the two component case. We assume constant
ellipticity and position angle. Dwarf galaxies, in reality, are not
necessarily smooth or have constant ellipticity. J+24 test an
alternative method using circular radii for the extended density
component, and we find these density profiles are very similar to the
fully elliptical case, even when assuming circular bins for the circular
outer component. As such, even reducing the assumed ellipticity from
\(0.37-0.55\) to 0 does not substantially impact the density profiles.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
\caption[Scl density comparison]{Density profiles for various
assumptions for Sculptor. PSAT is our fiducial 2-component J+24 sample,
circ is a 2-component bayesian model assuming circular radii, simple is
the series of simple cuts described, bright is the sample of the
brightest half of stars (scaled by 2), DELVE is a sample of RGB stars
(background subtracted and rescaled to
match).}\label{fig:scl_density_extras}
\end{figure}

Comparison of density profiles for each J+24 method. The fiducial is a
2-component elliptical model. However, the 1-component is still
elliptical but only contains 1 component and the circular model assumes
a circular outer density profile and bins in circular bins instead of
elliptical bins.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Scl literature density profiles]{A comparison of the Scl
derived density profile and historical works. Unlike most profiles in
this thesis, this density profile is plotted with respect to the
semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
The solid black line is a 2D exponential with corresponding scale radius
to Scl's half-light radius, and residuals in the lower panel are
relative to this exponential.}
\end{figure}

\subsection{Ursa Minor}\label{ursa-minor}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_extra.pdf}}
\caption[UMi density comparison]{Similar to
Fig.~\ref{fig:scl_observed_profiles} except for Ursa
Minor.}\label{fig:umi_density_extras}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_literature_profiles.pdf}}
\caption[UMi density profiles in literatures]{Similar to}
\end{figure}

\section{A comparison of density
profiles}\label{a-comparison-of-density-profiles}

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
histogram density profiles s compared to J+24. In general, both
methodologies agree well. However, J+24 tend to systematically
overestimate faint densities and confidently derive densities where the
MCMC model fails. Because J+24 use only one or two components across the
entire dwarf galaxy, the faint regions become sensitive to the spatial
likelihood. Antlia II is an extreme case, where the satellite is hidden
behind a large number of foreground stars. We derive a much more
poorly-constrained and fainter total density profile than J+24, and the
divergence between the MCMC method here and J+24's profile likely shows
that \emph{Gaia} observations are insufficient to properly constrain
Antia II's density profile. This is likely due to uncertainties in the
background density of satellite-like stars in Antlia II.

. Once the true density of stars drops below the background of
satellite-like stars, J+24's method will occasionally select consistent
background stars with a density directly dependent on the spatial
likelihood. The uncertainties in the outer regions are likely
underestimated.

@@

>>>>

In this section, we describe the details of J+24's likelihoods. We then
test the reliability of the derived density profiles for Scl and UMi,
comparing methodologies, samples, and to literature. Finally, we present
a comparison of J+24's samples to a non-parametric Bayesian density
profile, finding similar results in the inner regions. Past a ``limiting
radius'', J+24's density profiles become susceptible to the likelihood
specification. However, all methods agree up to the limiting radii. In
conclusion, the density profiles from J+24 are robust except in the very
outer regions.

@@

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\). Ensures high
  quality astrometry. \texttt{ruwe} is a measure of the excess
  astrometric noise on fitting a consistent parallax-proper motion
  solution \citep[see][]{lindegren+2021}.
\item
  3\(\sigma\) consistency of measured parallax with dwarf's distance,
  with \citet{lindegren+2021}'s zero-point correction. Note that the
  typical satellite parallax is very small.
\item
  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
  10\(\,{\rm mas\ yr^{-1}}\). This corresponds to tangental velocities
  of \(\gtrsim 500\) km/s at distances larger than 10 kpc.
\item
  Corrected colour excess is within 3\(\sigma\) of the expected
  distribution from \citet{riello+2021}. Removes stars with unreliable
  photometry.
\item
  De-reddened \(G\) magnitude is between
  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
  and stars significantly brighter than the tip of the red giant branch
  (TRGB) magnitude plus the distance modulus uncertainty
  \(\sigma_{\rm DM}\).
\item
  De-reddened colour is between \(-0.5 < {\rm BP - RP} <  2.5\). Removes
  stars substantially outside the expected CMD.
\end{itemize}

@@

The background likelihoods are instead empirically constructed. Stars
stars outside of 5\(R_h\) passing the quality cuts estimate the
background density in PM and CMD space. The density is a sum of
bivariate gaussians with variances based on Gaia uncertainties (and
covariance for proper motions). The spatial background likelihood is
assumed to be uniform over the field.

@@

\section{Alternative sample selection}\label{sec:simple_selection}

As J+24's probabilistic model may systematically bias the density
profile (see discussion below), we consider samples selected both from
\emph{Gaia} and external surveys using simple, absolute selection cuts.

For \emph{Gaia} data, we filter stars by four simple selection criteria.
We require \texttt{ruwe} \textless{} 1.3 as above, selecting only stars
with high-quality astrometry. We then require the star's parallax to be
\(3\sigma\)-consistent with the dwarf's distance. Next, we select stars
with proper motions with \(1\masyr\)\hspace{0pt} radius of the galaxy's
measured value. Finally, we filter stars to lie on the dwarf's CMD in
(\(G_{\rm BP} - G_{\rm RP}\),\(G\)), within the vertices defined in
Table~\ref{tbl:colour_cuts}. This polygon is derived based on the
colour-magnitude diagram of stars in the centre of the dwarf galaxy,
where the dwarf dominates the stellar density.

In addition to \emph{Gaia}, we select stars in the Scl field from the
deeper, photometric DELVE survey \citep{drlica-wagner+2022}. We select
sources DELVE data release 2 within an ellipse of radius 150 arcminutes.
We then only include sources which are categorized as likely stars and
sources with reliable (flags \textless=4) \(g\) and \(r\) magnitudes. We
then filter by the \(g\) and \(r\) magnitudes, selecting the empirical
CMD cut using the vertices in Table~\ref{tbl:colour_cuts}.

Ursa Minor is within the Ultraviolet Near-Infrared Optical Northern
Survey \citep[UNIONS][]{gwyn+2025}. We select sources within an ellipse
of radius 230 arminutes. We keep sources with no \texttt{FLAGS\_CFIS}
set and the \texttt{s21} and \texttt{s31} both \(<3\) (i.e.~not extended
sources). We then use the polygon bounded by the vertices in
Table~\ref{tbl:colour_cuts} to select the observed red giant branch,
horizontal branch, and main sequence of the dwarf.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/extra_cmd_selection.png}}
\caption[CMD cuts]{Colour-magnitude diagram cuts used in samples in this
section. Orange points are selected, black are non-selected stars within
\(1R_h\), and grey stars are for all stars. \textbf{todo, only plot
black/grey and cmd outline\ldots{}}}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/delve_unions_tangent.pdf}}
\caption[DELVE and UNIONS spatial distribution of stars]{The
distribution of DELVE and UNIONS selected stars in Scl and UMI}
\end{figure}

\section{\texorpdfstring{Uncertainties and possible biases in
\emph{Gaia}-derived density
profiles}{Uncertainties and possible biases in Gaia-derived density profiles}}\label{sec:density_extra}

In this section, we discuss additional tests and verification of the
derived density profiles. In particular, we check that alternative
methodologies do not substantially affect the density profile. In all
cases, the density profiles appear to have excellent convergence out to
\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
where the background dominates.

\textbf{Spatial likelihood.} J+24's algorithm was designed specifically
to detect the presence of a density excess in satellites and identify
individual stars for spectroscopic follow-up. We are instead interested
in constraining a dwarf galaxy's stellar structure. As such, one
potential problem with using J+24's candidate members is that the
stellar density model is either a single or double exponential. If the
(double) exponential model does not adequately describe a dwarf galaxy,
then J+24's derived density profiles may be systematically biased
towards the assumed (double) exponential profile.

As an example, in the lower panels of Fig.~\ref{fig:scl_density_extras};
Fig.~\ref{fig:umi_density_extras}, the density profile derived from the
fiducial (J+24) sample appears to be well-constrained at densities far
below the CMD+PM or \texttt{simple} model's background density. In
addition, the fiducial and 1-component density profiles deviate, despite
being derived identically besides the spatial likelihood assumption.

These stars are likely selecting stars from the statistical MW
background consistent either galaxy's CMD and PMs, recovering the
assumed density profile. As a result, the reliability of these density
profiles below the CMD+PM background may be questionable. A more robust
analysis, removing this particular density assumption, would be required
to more appropriately represent the knowledge of the density profile as
the background begins to dominate.

We revisit possible biases in spatial likelihood in
Section~\ref{sec:mcmc_hists}.

\textbf{Completeness}. \emph{Gaia} shows high but imperfect
completeness, particularly showing limitations in crowded fields and for
faint sources (\(G\gtrapprox20\)). In \citet{fabricius+2021}, for the
lowest density globular clusters, the completeness down to
\(G\approx 20\) is \(\sim 80\%\). As the lowest density globular
clusters are still far denser than dwarf galaxies, so the completeness
is likely higher for classical dwarfs. \emph{Gaia} is likely more
incomplete in terms of BP-RP photometry (as we use here) and for stars
with small pairwise separations (\(\lesssim 1.4''\)). The influence of
these details is unclear. Nevertheless, \emph{Gaia} should be more
complete for brighter stars.

We find little evidence for magnitude-dependent biases in the density
profiles. In
Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras} we show
density profiles derived using only stars brighter than the median
magnitude. We find no substantive differences compared to our fiducial
density profile. If \emph{Gaia}'s incompleteness is inhomogeneous, then
these variations do not appear to affect the density profiles.

If \emph{Gaia} has systematic density biases, the density profile should
vary among different facilities. We show the density profiles derived
from deep, ground-based UNIONS and DEVLE photometry (lower panels of
Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras}). For
both Scl and UMi, these alternative surveys result in near-identical
density profiles. Variations between density profiles derived from these
surveys appear to also be with uncertainties.

\textbf{Structural uncertainties}. The assumed structural parameters of
a dwarf may be misrepresentative or vary with radius. J+24 derive
probabilities where the outer stellar component is circular instead of
elliptical. Using the circular 2-component probabilities and binning in
circular radius, we find little difference in the derived density
profiles (the \texttt{circ} points in
Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras}). Thus,
our conclusions are robust to extreme variations in the ellipticity.
Reasonable changes in other structural parameters (centre, position
angle) are unlikely to have a more substantial effect.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/density_methods_extra.pdf}}
\caption[Sculptor density methodology comparison]{Density profiles for
various assumptions for Sculptor. \texttt{2-exp} is the fiducial
double-exponential-likelihood J+24 sample, \texttt{1-exp} instead is a
one-component exponential spatial likelihood, and \texttt{simple} uses
the simple position-independent selection criteria \texttt{circ} is a
2-component bayesian model assuming circular radii, simple is the series
of simple cuts described, bright is the sample of the brightest half of
stars (scaled by 2).}\label{fig:scl_density_extras}
\end{figure}

\section{A Bayesian density profile}\label{sec:mcmc_hists}

To address the concerns discussed above, we consider here a
non-parametric model to fit the density in each bin.

\subsection{Methodology}\label{methodology}

This model extends the J+24 framework with two notable differences. (1)
The systemic proper motions are held fixed. This allows for a much more
efficient calculation of the likelihoods. (2) The spatial likelihood for
the satellite is a piecewise constant density function.While this model
theoretically has many more parameters to fit (one for each bin), each
parameter is independently estimated based on only the stars in that
specific bin.

@@

In detail, we parameterize \(f_{\rm sat}\) in terms of the log-relative
satellite density. The model construction is then \begin{equation}{
\begin{align}
\theta_i &\equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}}) \\
f_{\rm sat} &= \frac{10^{\theta_i}}{1 + 10^{\theta_i}} \\
\theta_i &\sim {\rm Uniform}(-12, 6) \\
{\cal L}_{\rm tot} &= f_{\rm sat}{\cal L}_{\rm CMD, PM} + (1-f_{\rm sat}){\cal L}_{\rm CMD, PM} \\
\end{align}
}\end{equation}

For each satellite, we hold the structural parameters fixed to those in
J+24, create bins which are the larger of the bin containing 20 stars or
0.05dex in \(\log R_{\rm ell}\). We then sample each chain 1000 steps
for 48 walkers using Turing.jl and the No-U Turns Sampler.

@@

\subsection{Results}\label{results}

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
histogram density profiles compared to J+24. In general, both
methodologies agree well. However, J+24 tend to systematically
overestimate faint densities and confidently derive densities where the
MCMC model fails. Because J+24 use only one or two components across the
entire dwarf galaxy, the faint regions become sensitive to the spatial
likelihood. Antlia II is an extreme case, where the satellite is hidden
behind a large number of foreground stars. We derive a much more
poorly-constrained and fainter total density profile than J+24, and the
divergence between the MCMC method here and J+24's profile likely shows
that \emph{Gaia} observations are insufficient to properly constrain
Antia II's density profile. This is likely due to uncertainties in the
background density of satellite-like stars in Antlia II.

Once the true density of stars drops below the background of
satellite-like stars, J+24's method will occasionally select consistent
background stars with a density directly dependent on the spatial
likelihood. The uncertainties in the outer regions are likely
underestimated.

@@

\section{Comparison to literature}\label{comparison-to-literature}

We finally compare our density profiles against a sample of
literature-derived density profiles in
Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. The
J+24-derived density profiles are representative of state-of-the-art
density profiles. A few photometric surveys may extend to slightly
fainter densities, but typically with large uncertanties. Deviations
between sources are slight despite the range of methods across time. All
density profiles extending into the outskirts of Scl and UMi show a
similar overdensity to the ones we find. The extended stellar profiles
of Scl and UMi appear to be a robust result across the literature.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Sculptor literature density profiles]{A comparison of the Scl
derived density profile and historical works. Unlike most profiles in
this thesis, this density profile is plotted with respect to the
semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
The solid black line is a 2D exponential with corresponding scale radius
to Scl's half-light radius, and the residuals in the bottom panel are
with respect to this profile. \textbf{remove delve, umi title\ldots,,
combine??}}\label{fig:scl_lit_profiles}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_literature_profiles.pdf}}
\caption[Ursa Minor literature density profiles]{Similar to
Fig.~\ref{fig:scl_lit_profiles} except for Ursa Minor. The references
are (in order) \citet{sato+2025}; \citet{palma+2003};
\citet{martinez-delgado+2001}; \citet{kleyna+1998}; \citet{IH1995}; and
\citet{Hodge1964}.}
\end{figure}

\section{Summary}\label{summary}

In this Appendix, we discussed the methodological details of the J+24
sample selection algorithm. We consider possible biases due to the
assumed spatial likelihood, \emph{Gaia}'s completeness, and structural
parameters. In all cases, we find these assumptions likely do not cause
major biases in the derived density profiles. However, we show that
J+24's density profiles become unreliable when dropping below the
background of satellite-like stars. We address this concern through an
algorithm rederiving the density in elliptical bins. We find excellent
agreement with J+24 for most galaxies out to a background-limited
radius. Finally, comparing our density profiles against the literature,
we find our results to be consistent. We conclude that the detection of
an extended density profile in Scl and UMi is robust to incompleteness,
methodology, alternative surveys, and across the literature.

### 2025-10-06 thesis/extra_numerical_convergence.tex
thesis; julio oct 4 comments

Here, we describe some convergence tests to ensure our methods and
results are minimally impacted by numerical limitations and assumptions.
See \citet{power+2003} for a detailed discussion of various assumptions
and parameters used in N-body simulations.

\section{Softening}\label{softening}

One challenge of N-body integration is close, collisional encounters,
assuming point particles, cause divergences in the local force. However,
this should not occur in a \emph{collisionless} simulation. As a result,
most collisionless N-body codes adopt a gravitational softening, a
length scale below which the force of gravity begins to decrease between
point particles.

\citet{power+2003} empirically suggest that the ideal softening is
\begin{equation}{
h_{\rm grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}},
}\end{equation}

where \(h_{\rm grav}\) is the softening length, and \(N_{200}\) is the
number of particles within \(R_{200}\). This choice balances integration
time and only compromises resolution in collisional regime.

For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
slightly smaller softening which was reduced by a factor of
\(\sqrt{10}\) which appears to improve agreement slightly in the
innermost regions.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_softening.png}}
\caption{Softening convergence}\label{fig:softening_convergence}
\end{figure}

\section{Time stepping and force
accuracy}\label{time-stepping-and-force-accuracy}

In general, we use adaptive timestepping and relative opening criteria
for gravitational force computations. To verify that these choices and
associated accuracy parameters minimally impact convergence or speed, we
show a few more isolation runs (using only 1e5 particles)

\begin{itemize}
\tightlist
\item
  constant timestep (\ldots), approximantly half of minimum timestep
  with adaptive timestepping
\item
  geometric opening, with \(\theta = 0.5\).
\item
  strict integration accuracy, (facc = \ldots.)
\end{itemize}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_methods.png}}
\caption{Isolation method convergence}\label{fig:methods_convergence}
\end{figure}

>>>>

Here, we describe the numerical specifics of our simulation and
analysis. We present convergence tests supporting our choices of
softening and additional parameters. We find that alternate choices of
numerical parameters neither improve convergence nor the resulting
evolution.

Due to the finite resolution of N-body simulations, collisional (close)
encounters between particles are inevitable. Such collisional encounters
are both unphysical (with possible arbitrary acceleration, violating the
``collisionless'' assumption) and computationally expensive (requiring
many small time-steps). As a remedy, many N-body codes use a
``softened'' gravitational force law, where the force weakens when
closer to a particle than a ``softening length''. The choice of
softening length ideally balances resolution and computational speed.

Empirically, \citet{power+2003} suggest that the ideal softening is
\begin{equation}\protect\phantomsection\label{eq:power_softening}{
h_{\rm grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}},
}\end{equation}

where \(h_{\rm grav}\) is the softening length, and \(N_{200}\) is the
number of particles within \(R_{200}\). This choice balances integration
time and only compromises resolution in collisional regime. For our
isolation halo (\(\rmax=6\,\kpc\), \(\vmax=31\,\kms\)) with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\). Next we consider
the effect of softening as applied to our isolation halo.

The two other key numerical parameters in \gadget{} are the tree-force
and time-step accuracy parameters. Specifically, \gadget{} opens a node
if \(M\,l/r^3 < \alpha |a|\), the node's mass is \(M\), distance from
the particle is \(r\), side-length is \(l\), and the particle's total
acceleration is \(a\). We adopt \(\alpha =0.005\). We also elect to use
adaptive time stepping with integration accuracy set to \(\eta=0.01\).
(particles must take time-steps smaller than
\(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)).

The top panel of Fig.~\ref{fig:methods_convergence} illustrates the
influence of softening length for simulations of Scl's \smallperi{}
orbit with \(10^5\) particles. We show a model with the fiducial
softening length (\(h=0.14\,\kpc\)), and larger and smaller softenings
by a factor \(\sqrt{10}\). The larger softening length is consistent
with Eq.~\ref{eq:power_softening}'s prediction. With a larger softening
length, the halo diverges from the expectation (using \(100\times\) more
particles) across most radii. On the other hand, the fiducial and
smaller softening lengths predict similar final results, although the
smaller softening length simulation deviates slightly more than the
fiducial, likely a result of more collisional encounters. As computation
time increases moderately with decreasing softening length, the fiducial
softening length balances efficiency and accuracy for this simulation.
While smaller than \citet{power+2003}'s suggestion, our choice is
designed for more compact, dwarf satellite halos instead of large,
galaxy-mass isolated halos.

The lower panel of Fig.~\ref{fig:methods_convergence} tests changes to
these parameters. We include a model with smaller integration accuracy
(``small timestep'') and a stricter tree-force tolerance (``high acc.
force''). More precise tolerances on these parameters does not affect
the evolution within the uncertainties of the final profile.

As demonstrated in this subsection, stricter numerical accuracy and
modified softening lengths do not affect our results. We thus conclude
that our simulations are numerically well-converged (up to the
``convergence'' radius).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/orbit_converg_methods.png}}
\caption[Isolation method convergence]{A comparison of the final
profiles using different simulation methods for Sculptor's \smallperi{}
model. The benchmark model is our fiducial \(10^7\) particle run, and
all other models use \(10^5\) paritcles, with their ``converged radius''
marked by the black arrow and softening by the vertical bar.
\textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
lengths. \textbf{Bottom:} Models with more precise timestep accuracy
(\(\eta=0.001\) compared to fiducial \(\eta=0.005\)) and gravitational
force accuracy (\(\alpha=0.003\) versus \(\alpha=0.01\)).
\textbf{Summary:} Except for the model with a larger softening length,
all simulations agree within uncertainties. \textbf{Todo: softening
along lower axis always\ldots, changing
linewidth?}}\label{fig:methods_convergence}
\end{figure}
